---
title: "survOmicsPkg"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE,
  comment = "#>")
```


survOmicsPkg provides tools for preparing longitudinal data for survival analysis. It supports construction of metadata for baseline biomarker and time varying biomarker analyses and flexible survival analyses on omics datasets. 

Install the package from Github:

```{r}

#install.packages("remotes")
library("remotes")
remotes::install_github("lingdi-zhang/survOmicsPkg")

```
Overview

This package is designed for workflows involving:

longitudinal or repeated-measures data
survival or time-to-event analysis
biomarker discovery pipelines
reproducible preprocessing of omics or clinical datasets


Load required packages:
```{r }
library(survOmicsPkg)

```
Input Data Format

The package expects a long-format metadata dataset where each row represents one subject at one time point.

```{r example for baseline}
toy_metadata <- data.frame(
  id = c(1,1,1,2,2,3,3),
  time = c(0,2,6,0,2,0,2),
  event = c(0,0,1,0,1,0,0))
toy_metadata
```
Required columns for metadata

id: subject identifier
time: time variable (numeric)
event: event indicator (0/1)


For baseline analysis
The package expects a baseline biomarker dataframe and a vector of biomarker feature names 
```{r example for baseline}
baseline_biomarkers<-data.frame(
  id=c(1,2,3),
  biomarker1=c(10.2,8.5,7.2),
  biomarker2=c(5.1,4.8,6.5)
)
baseline_biomarkers

biomarkers=c("biomarker1","biomarker2")
biomarkers

```
Required columns for baseline biomarker dataset

id: subject identifier
feature columns: biomarkers, genes, or other measurements

Workflow

A typical workflow:

Prepare longitudinal data in long format
Construct datasets using preprocess_data()
run run_multiple_cox_flexible() for omics data with multiple biomarkers
run calculate_FDR for compute FDR for each effect type (main effect, time effect, covariates or interactions term) from the results 

```{r construct metadata for baseline biomarker analysis}
processed_data<-preproccess_data(
  metadata=toy_metadata,biomarkers=biomarkers,id_col="id",event_col="event",time_col="time",baseline_feature_table=baseline_biomarkers,biomarker_type="baseline")

processed_data

res<-run_multiple_cox_flexible(processed_data,biomarkers=biomarkers, id_col="id",event_col="event",time_col="time",biomarker_type="baseline")

outcomes=res[[1]]
terms=res[[2]]
FDR_res<-calculate_FDR(outcomes,terms)

head(FDR_res)
```
For baseline analysis with time varying outcomes
set time_varying_coefficients=TRUE and adjust center_time if needed

```{r}
res<-run_multiple_cox_flexible(processed_data,biomarkers=biomarkers, id_col="id",event_col="event",time_col="time",biomarker_type="baseline",time_varying_coefficients=TRUE,center_time=0)

outcomes=res[[1]]
terms=res[[2]]
FDR_res<-calculate_FDR(outcomes,terms)

head(FDR_res)
```

For time varying biomarker analysis
The package expects a time varying biomarker dataframe and a vector of biomarker feature names 

```{r time varying biomarkers}
time_varying_biomarkers<-data.frame(
  id = c(1,1,1,2,2,3,3),
  time = c(0,2,6,0,2,0,2),
  biomarker1 = c(10.2, 11.4, 13.1, 8.5, 9.7, 7.2, 7.8),
  biomarker2 = c(5.1, 5.5, 6.2, 4.8, 5.0, 6.5, 6.3)
)

time_varying_biomarkers
biomarkers=c(biomarker1,biomarker2)
biomarkers

```

Required columns for time varying biomarker dataset

id: subject identifier
time: time variable (numeric)
feature columns: biomarkers, genes, or other measurements

Workflow

A typical workflow:

Prepare longitudinal data in long format
Construct datasets using preprocess_data()
run run_multiple_cox_flexible() for omics data with multiple biomarkers
run calculate_FDR for compute FDR for each effect type (main effect, time effect, covariates or interactions term) from the results 

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
```{r}
processed_data<-preproccess_data(
  metadata=toy_metadata,biomarkers=biomarkers,id_col="id",event_col="event",time_col="time",start_col= 'start', stop_col='stop',time_varying_feature_table=time_varying_biomarkers,biomarker_type="time_varying")

processed_data

res<-run_multiple_cox_flexible(processed_data,biomarkers=biomarkers, id_col="id",event_col="event",start_col= 'start', stop_col='stop',biomarker_type="time_varying")

outcomes=res[[1]]
terms=res[[2]]
FDR_res<-calculate_FDR(outcomes,terms)

head(FDR_res)
```

Run analysis with baseline biomarker and (time varying biomarker-baseline biomarker)

```{r}
processed_data<-preproccess_data(
  metadata=toy_metadata,biomarkers=biomarkers,id_col="id",event_col="event",time_col="time",start_col= 'start', stop_col='stop',baseline_feature_table=baseline_biomarkers,time_varying_feature_table=time_varying_biomarkers,biomarker_type="baseline_change")

processed_data

res<-run_multiple_cox_flexible(processed_data,biomarkers=biomarkers, id_col="id",event_col="event",start_col= 'start', stop_col='stop',biomarker_type="baseline_change")

outcomes=res[[1]]
terms=res[[2]]
FDR_res<-calculate_FDR(outcomes,terms)

head(FDR_res)
```

