---
title: "survOmicsPkg"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE,
  comment = "#>")
```


survOmicsPkg provides tools for preparing longitudinal data for survival analysis. It supports construction of metadata for baseline biomarker and time varying biomarker analyses and flexible survival analyses on omics datasets. 

Install the package from Github:

```{r}
library("remotes")
remotes::install_github("lingdi-zhang/survOmicsPkg", force=TRUE)

```
Overview  

This package is designed for workflows involving:

longitudinal or repeated-measures data  
survival or time-to-event analysis  
reproducible preprocessing of omics or clinical datasets  

A typical workflow:  

Prepare longitudinal data in long format  
Construct datasets using preprocess_data()  
run run_multiple_cox_flexible() for omics data with multiple biomarkers  
run calculate_FDR for compute FDR for each effect type (main effect, time effect, covariates or interactions term) from the results   


Load required packages:
```{r}
library(survOmicsPkg)

```
Input Metadata Format  

The package expects a long-format metadata dataset where each row represents one subject at one time point.

```{r}
toy_metadata <- data.frame(
  id = c(1,1,1,2,2,3,3),
  time = c(0,2,6,0,2,0,2),
  event = c(0,0,1,0,1,0,0))
toy_metadata
```
Required columns for metadata

id: subject identifier  
time: time variable (numeric)  
event: event indicator (0/1)  

For baseline biomarker analysis: 

The package expects a baseline biomarker dataframe and a vector of biomarker feature names 
```{r}
baseline_biomarkers<-data.frame(
  id=c(1,2,3),
  biomarker1=c(10.2,8.5,7.2),
  biomarker2=c(5.1,4.8,6.5)
)
baseline_biomarkers

biomarkers=c("biomarker1","biomarker2")
biomarkers

```
Required columns for baseline biomarker dataset

id: subject identifier  
feature columns: biomarkers, genes, or other measurements

This analysis test for: 
h(t∣x)=h0(t)exp(βx)  
β is biomarker effect on log hazard ratio

```{r}
processed_data<-preproccess_data(
  metadata=toy_metadata,biomarkers=biomarkers,id_col="id",event_col="event",time_col="time",baseline_feature_table=baseline_biomarkers,biomarker_type="baseline")

processed_data

res<-run_multiple_cox_flexible(processed_data,biomarkers=biomarkers, event_col="event",time_col="time",biomarker_type="baseline")

outcomes=res[[1]]
terms=res[[2]]
FDR_res<-calculate_FDR(outcomes,terms)

head(FDR_res)
```
For baseline analysis with time varying outcomes: 

set time_varying_coefficients=TRUE and adjust center_time if needed

This analysis test for: 
h(t∣x)=h0(t)exp(β1x+β2x⋅t)  
β1 is biomarker effect on log hazard ratio at t=0  
β2 is the interaction effect: how the main effect change over time  

This is an example, with the toy data, the interaction term won't converge. 
```{r}
res<-run_multiple_cox_flexible(processed_data,biomarkers=biomarkers, event_col="event",time_col="time",biomarker_type="baseline",time_varying_coefficients=TRUE,center_time=0)

outcomes=res[[1]]
terms=res[[2]]
FDR_res<-calculate_FDR(outcomes,terms)

head(FDR_res)
```

For time varying biomarker analysis: 
The package expects a time varying biomarker dataframe and a vector of biomarker feature names 

This analysis tests for:
h(t∣x(t))=h0(t)exp(β⋅x(t))  
β is biomarker effect on log hazard ratio, while the biomarker values changes over time  

```{r}
time_varying_biomarkers<-data.frame(
  id = c(1,1,1,2,2,3,3),
  time = c(0,2,6,0,2,0,2),
  biomarker1 = c(10.2, 11.4, 13.1, 8.5, 9.7, 7.2, 7.8),
  biomarker2 = c(5.1, 5.5, 6.2, 4.8, 5.0, 6.5, 6.3)
)

time_varying_biomarkers
biomarkers=c("biomarker1","biomarker2")
biomarkers

```
Required columns for time varying biomarker dataset

id: subject identifier  
time: time variable (numeric)  
feature columns: biomarkers, genes, or other measurements  



```{r}
processed_data<-preproccess_data(
  metadata=toy_metadata,biomarkers=biomarkers,id_col="id",event_col="event",time_col="time",start_col= 'start', stop_col='stop',time_varying_feature_table=time_varying_biomarkers,biomarker_type="time_varying")

processed_data

res<-run_multiple_cox_flexible(processed_data,biomarkers=biomarkers, event_col="event",start_col= 'start', stop_col='stop',biomarker_type="time_varying")

outcomes=res[[1]]
terms=res[[2]]
FDR_res<-calculate_FDR(outcomes,terms)

head(FDR_res)
```

For analysis consider both baseline biomarker effect and within-subject change from baseline: 

This analysis tests for:
h(t)=h0(t)exp(β1x0+β2Δx(t))
β1 represents the baseline biomarker effect on log hazard  
β2 represents the effect of within-subject change from baseline on log hazard  


This is an exmaple, with the toy dataset, the β2 term won't converge. 
```{r}
processed_data<-preproccess_data(
  metadata=toy_metadata,biomarkers=biomarkers,id_col="id",event_col="event",time_col="time",start_col= 'start', stop_col='stop',baseline_feature_table=baseline_biomarkers,time_varying_feature_table=time_varying_biomarkers,biomarker_type="baseline_change")

processed_data

res<-run_multiple_cox_flexible(processed_data,biomarkers=biomarkers, event_col="event",start_col= 'start', stop_col='stop',biomarker_type="baseline_change")

outcomes=res[[1]]
terms=res[[2]]
FDR_res<-calculate_FDR(outcomes,terms)

head(FDR_res)
```

